---
title: "When Gender Matters"
author: "Audrey Loritz"
description: "Party Differences in Female Candidate Success in Primary Elections"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(infer)
```

```{r, include=FALSE}

dem_candidates <- read_csv("data/dem_candidates.csv")
rep_candidates <- read_csv("data/rep_candidates.csv")
```


## Introduction

**How does gender impact candidate success in primary elections for the different political parties?**

Gender has long been considered an important factor in electoral politics, particularly in recent decades as the number of female candidates has increased across party lines. By narrowing in on primary elections, it is possible to examine gendered differences in primary success both within and between the Democratic and Republican parties. Understanding this relationship is important as it sheds light on whether barriers or advantages for women exist and whether these patterns differ across partisan contexts.

I hypothesize that the effect of gender on candidate success will differ by party. Specifically, I expect that female candidates will perform better in Democratic primaries than in Republican primaries, where gender may have little or even negative effects. This assumption stems from the idea that Democratic electorates tend to be more supportive of female candidates, while Republican electorates may be less responsive to gender as an advantage. As a result, gender is likely to have a conditional effect on electoral success rather than a uniform one across parties.

## Data

```{r, include=FALSE}

elections <- bind_rows(dem_candidates, rep_candidates,
                       .id = "Party")

elections <- elections |>
  select(Candidate, Party, Gender, `Race 1`, Office, `Primary Outcome`, `Primary %`, `Runoff Outcome`, `Runoff %`) |>
   mutate(Party = if_else(Party == 1, "Democrat", "Republican"))

elections <- elections |>
  mutate(`Primary %` = parse_number(`Primary %`)) |>
  filter(Gender != "Nonbinary")
```

The explanatory variables in this paper are party, determined by which primary the candidate is running in, and gender, determined by campaign materials and interviews with candidates. The outcome variable is primary election success, which is measured by the percentage of votes a candidate got during the 2022 primary election. 


The graph below displays the distribution of primary election vote share, the main outcome of interest. Additionally, it indicates the different distribution of male and female vote shares to illustrate how that variation of interest looks.

```{r, echo=FALSE}

interest <- ggplot(elections, aes(x = `Primary %`, fill = Gender)) +
  geom_histogram() +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  labs(x = "Average Primary Vote Percentage", y = "Count",
       title = "Distrabution of Primary Vote Percentages by Gender",
       subtitle = "Data from 2022 primary elections")

interest
```
For this paper, there are two comparisons of interest: the difference between male and female candidates within each party, and the difference in effect gender has on the Democratic and Republican parties. To examine this, I will utilize the two hypotheses listed below.

**Party Hypothesis**

*Null Hypothesis:* Gender will have no impact on primary election results in either party.

*Alternative Hypothesis:* Gender affects primary election success; this effect may differ across parties.

Using α = 0.05, the null hypothesis can be rejected if the p-value for one or both of the political parties is less than 0.05. This would indicate that gender does affect primary election results, even if that effect varies across party lines.

**Interaction Hypothesis**

*Null Hypothesis:* The effect of gender on primary election success will not vary by political party.

*Alternative Hypothesis:* The effect of gender on primary election success will vary by political party.

Using α = 0.05, the null hypothesis can be rejected if the p-value of the interaction coefficient is less than 0.05. This would indicate that there is a difference between the role that gender plays in Democratic and Republican primary elections.

## Results

```{r, echo=FALSE}

party_gender_success <- elections |> group_by(Party, Gender) |>
           summarise(avg_primary = mean(`Primary %`, na.rm = TRUE)) |>
  ggplot(aes(x = Party, y = avg_primary, fill = Gender)) + 
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  labs(x = "Party", y = "Average Primary Vote Percentage", 
       title = "Gender Breakdown of Primary Success by Party", 
       subtitle = "Data from 2022 primary elections")

party_gender_success
```

This graph demonstrates primary success based on gender within the two main political parties. The average Democratic vote percentage for female candidates is just over 50%, whereas female Republicans only got an average of under 30% in their elections. Female candidates in Democratic primaries saw greater success than those in Republican primaries.



```{r, echo=FALSE}

fit <- lm(`Primary %` ~ 
            relevel(as.factor(Gender), ref = "Male") * 
            relevel(as.factor(Party), ref = "Republican"),
          data = elections)

fit_table <- tidy(fit) |>
  select(term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate)

knitr::kable(fit_table, digits = 2,
             col.names = c("Intercept(Male Republicans)", "Female Republicans",
                           "Male Democats", "Female Democrats"))
```




*Standard Error*

```{r, echo=FALSE}

std_er <- tidy(fit) |>
  select(term, std.error) |>
  pivot_wider(names_from = term, values_from = std.error)

knitr::kable(std_er, digits = 2,
             col.names = c("Intercept(Male Republicans)", "Female Republicans",
                           "Male Democats", "Female Democrats"))
```

*P-Values*

```{r, echo=FALSE}

p_values <- tidy(fit) |>
  select(term, p.value) |>
  pivot_wider(names_from = term, values_from = p.value)

knitr::kable(p_values, digits = 2,
             col.names = c("Intercept(Male Republicans)", "Female Republicans",
                           "Male Democats", "Female Democrats"))
```

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```
