---
title: "When Gender Matters"
author: "Audrey Loritz"
description: "Party Differences in Female Candidate Success in Primary Elections"
output:
  distill::distill_article:
    self_contained: false
---

```{r setup, include=FALSE}
library(tidyverse)
library(broom)
library(infer)
```

```{r, include=FALSE}

dem_candidates <- read_csv("data/dem_candidates.csv")
rep_candidates <- read_csv("data/rep_candidates.csv")
```


## Introduction

**How does gender impact candidate success in primary elections for the different political parties?**

Gender has long been considered an important factor in electoral politics, particularly in recent decades as the number of female candidates has increased across party lines. By narrowing in on primary elections, this analysis not only looks at gendered differences but also at how those differences vary across the partisan divide. Making it possible for this paper to examine gendered differences in primary success both within and between the Democratic and Republican parties. Understanding this relationship is important as it sheds light on whether barriers or advantages for women exist and whether these patterns differ across partisan contexts.

I hypothesize that the effect of gender on candidate success will differ by party. Specifically, I expect that female candidates will perform better in Democratic primaries than in Republican primaries, where gender may have little or even negative effects. This assumption stems from the idea that Democratic electorates tend to be more supportive of female candidates, while Republican electorates may be less responsive to gender as an advantage. As a result, gender is likely to have a conditional effect on electoral success rather than a uniform one across parties.

## Data

```{r, include=FALSE}

elections <- bind_rows(dem_candidates, rep_candidates,
                       .id = "Party")

elections <- elections |>
  select(Candidate, Party, Gender, `Race 1`, Office, `Primary Outcome`, `Primary %`, `Runoff Outcome`, `Runoff %`) |>
   mutate(Party = if_else(Party == 1, "Democrat", "Republican"))

elections <- elections |>
  mutate(`Primary %` = parse_number(`Primary %`)) |>
  filter(Gender != "Nonbinary")
```


```{r, echo=FALSE}

interest <- ggplot(elections, aes(x = `Primary %`, fill = Gender)) +
  geom_histogram() +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  labs(x = "Average Primary Vote Percentage", y = "Count",
       title = "Distrabution of Primary Vote Percentages by Gender",
       subtitle = "Data from 2022 primary elections")

interest
```


## Results

```{r, echo=FALSE}

party_gender_success <- elections |> group_by(Party, Gender) |>
           summarise(avg_primary = mean(`Primary %`, na.rm = TRUE)) |>
  ggplot(aes(x = Party, y = avg_primary, fill = Gender)) + 
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("Female" = "pink", "Male" = "lightblue")) +
  labs(x = "Party", y = "Average Primary Vote Percentage", 
       title = "Gender Breakdown of Primary Success by Party", 
       subtitle = "Data from 2022 primary elections")

party_gender_success
```

This graph demonstrates primary success based on gender within the two main political parties. The average Democratic vote percentage for female candidates is just over 50%, whereas female Republicans only got an average of under 30% in their elections. Female candidates in Democratic primaries saw greater success than those in Republican primaries.



```{r, echo=FALSE}

fit <- lm(`Primary %` ~ 
            relevel(as.factor(Gender), ref = "Male") * 
            relevel(as.factor(Party), ref = "Republican"),
          data = elections)

fit_table <- tidy(fit) |>
  select(term, estimate) |>
  pivot_wider(names_from = term, values_from = estimate)

knitr::kable(fit_table, digits = 2,
             col.names = c("Intercept(Male Republicans)", "Female Republicans",
                           "Male Democats", "Female Democrats"))
```



Standard Error

```{r, echo=FALSE}

std_er <- tidy(fit) |>
  select(term, std.error) |>
  pivot_wider(names_from = term, values_from = std.error)

knitr::kable(std_er, digits = 2,
             col.names = c("Intercept(Male Republicans)", "Female Republicans",
                           "Male Democats", "Female Democrats"))
```

P-Values

```{r, echo=FALSE}

p_values <- tidy(fit) |>
  select(term, p.value) |>
  pivot_wider(names_from = term, values_from = p.value)

knitr::kable(p_values, digits = 2,
             col.names = c("Intercept(Male Republicans)", "Female Republicans",
                           "Male Democats", "Female Democrats"))
```

```{r print-code, ref.label=knitr::all_labels(), echo = TRUE, eval=FALSE}

```
